A recipe for extremely reproducible enrichment analysis (and why we need it)

### 2023-12-08 manuscript/manuscript.Rmd
data

Unreliable and irreproducible research is a significant problem that wastes resources and risks
undermining the public perception of science.
Previous work has highlighted that published enrichment analyses frequently suffer from statistical
and reporting flaws.
We sought to determine whether this translates into irreproducibility by examining whether the
findings of 20 open-access articles published in 2019 describing enrichment analysis with the
popular DAVID suite were reproducible.
We find that only three articles exhibited a high degree of concordance, while seven exhibited major
discrepancies, which we mainly ascribe to deficiencies in methodological reporting.
As the tool version used is no longer available, all articles using this tool pre-2021 (~20,800 studies)
including this sample of 20 articles are no longer able to be reproduced with the original tools.
Based on this, we suggest that results from web-based tools without long-term preservation features should
not be included in scientific publications due to the threat of link decay and short reproducibility horizon.
Relying exclusively on webtools for analysis may also be in breach of institutional and funder data management mandates.
We advocate for the adoption of extremely reproducible research workflows, and we provide a detailed protocol
for how to achieve it using a combination of best practices including literate programming, version control,
containerisation, documentation and persistent sharing of data and software.

@@

library("DiagrammeR")
library("RColorBrewer")

@@

@@

Briefly, we used the DAVID "Analysis Wizard" and pasted in the provided gene list into the text box,
selecting this as "gene list", together with the appropriate gene identifier type.
Typically this was the official gene symbol.
The species was specified as *Homo sapiens*.
Next we clicked on the "Functional Annotation Tool" link, bringing up the "Annotation Summary Results”
page.
We followed the steps indicated in the respective papers to select appropriate gene sets investigated
in the articles such as Gene Ontology (GO; biological process, molecular function), Kyoto Encyclopedia
of Genes and Genomes (KEGG) pathways, etc.
Next, the "Functional Annotation Chart" was clicked on, opening a new window with a table of
enrichment results, which were saved for future reference.

@@

After conducting this reproduction study according to the methods described in the respective articles,
the 20 articles were classified based on the degree of reproducibility (**Figure 1**).
Seven studies were classified as low reproducibility, while 10 were classified as medium and just three
were classified as high.
Justifications for these classifications are provided in Table S1.

@@

Although this reproducibility study is relatively small, there are a few key observations.
We observe that only 3/20(15%) of studies achieved a high degree of reproducibility.
In a way, this is unexpected, given the simplicity of using DAVID.
Users provide a list of gene idetifiers in a text box and select some options.
There are several posible explanations for this inconsistency:

@@

@@

This article is an attempt to foster increased reproducibility in genomics by demonstrating the
implementation of the five pillars principles into a commonly used routine, namely differential
expression and pathway enrichment analysis.
However, these concepts may be difficult for novice analysts, and so we provide a step-by-step
protocol to guide them on how to customise template Dockerfile and R Markdown scripts to suit their
project's needs [@Ziemann2023-kz].
Although we aimed to make this process as easy as possible, it requires a substantial time outlay, which
we estimate at 15-20 hours.
This is significantly longer than using a web-based tool that provides results in just a few
minutes, but we argue the time investment is well spent.
It ensures analytical reproducibility, assists with auditing and makes the process of reproduction
simple and fast (Box 1), which should aid the early identification of errors.

>>>>

Unreliable and irreproducible research is a significant problem that wastes resources and risks
undermining the public perception of science.
Previous work has highlighted that published enrichment analyses frequently suffer from statistical
and reporting flaws.
We sought to determine whether this translates into irreproducibility by examining whether the
findings of 20 open-access articles published in 2019 describing enrichment analysis with the
popular DAVID suite were reproducible.
We find that only four articles exhibited a high degree of concordance, while seven exhibited major
discrepancies, which we mainly ascribe to deficiencies in methodological reporting.
As the tool version used is no longer available, all articles using this tool pre-2021 (~20,800 studies)
including this sample of 20 articles are no longer able to be reproduced with the original tools.
Based on this, we suggest that results from web-based tools without long-term preservation features should
not be included in scientific publications due to the threat of link decay and short reproducibility horizon.
Relying exclusively on webtools for analysis may also be in breach of institutional and funder data management mandates.
We advocate for the adoption of extremely reproducible research workflows, and we provide a detailed protocol
for how to achieve it using a combination of best practices including literate programming, version control,
containerisation, documentation and persistent sharing of data and software.

@@

library("DiagrammeR")
library("RColorBrewer")
#library("kableExtra") there is a bug with kableextra - waiting for patch
library("gt")

@@

We screened reporting criteria of these articles, including whether the version of the application
is given, whether p-values underwent correction for multiple testing, whether a background list was
described and whether gene selection was described unambiguously.

@@

Briefly, we used the DAVID v6.8 "Analysis Wizard" and pasted in the provided gene list into the text box,
selecting this as "gene list", together with the appropriate gene identifier type.
Typically this was the official gene symbol.
The species was specified as *Homo sapiens*.
Next we clicked on the "Functional Annotation Tool" link, bringing up the "Annotation Summary Results”
page.
We followed the steps indicated in the respective papers to select appropriate gene sets investigated
in the articles such as Gene Ontology (GO; biological process, molecular function), Kyoto Encyclopedia
of Genes and Genomes (KEGG) pathways, etc.
Next, the "Functional Annotation Chart" was clicked on, opening a new window with a table of
enrichment results, which were saved for future reference.

@@

As reproducibility is associated with the quality of documentation we collected key methodological information
about how the enrichment analysis was done and present in Table 1.
From this, we can see that only seven articles reported the version of DAVID that was used.
Version 6.7 was reportedly used in two articles while version 6.8 was specified in five.
Correction of p-values for multiple testing was specified or inferred in 11 articles, while the results of
the remaining nine articles did not take into consideration the importance of p-value correction for
minimising false positives.
Of the 20 articles, none mentioned explicitly that a background list was used for enrichment analysis.
We also examined whether the gene selection criteria was unambiguously clear, which was true in 16 articles.

```{r,table1,echo=FALSE}

x <- read.table("../data/checklist.tsv",header=TRUE,row.names=1)

colnames(x) <- gsub("_"," ",colnames(x))
x |> gt(rownames_to_stub = TRUE) |> tab_header(title = "Table 1. Article reporting checklist results.")
#x %>% kbl(caption = "Table 1. Article reporting checklist results.") %>% kable_paper("hover", full_width = F)

We attempted reproduction according to the methods described in the respective articles, with the limitation
that we had to use DAVID v6.8 as v6.7 was no longer available.
The 20 articles were classified based on the degree of reproducibility (**Figure 1**).
Seven studies were classified as low reproducibility, while 9 were classified as medium and just four
were classified as high.
Justifications for these classifications are provided in Table S1.

@@

Although this reproducibility study is relatively small, there are a few key observations.
We observe that only 4/20(20%) of studies achieved a high degree of reproducibility.
In a way, this is unexpected, given the simplicity of using DAVID.
Users provide a list of gene idetifiers in a text box and select some options.
There are several posible explanations for this inconsistency:

@@

* Differences between DAVID version 6.7 and 6.8.

@@

This article is an attempt to foster increased reproducibility in genomics by demonstrating the
implementation of the five pillars principles into a commonly used routine, namely differential
expression and pathway enrichment analysis.
However, these concepts may be difficult for novice analysts, and so we provide a step-by-step
protocol to guide them on how to customise template Dockerfile and R Markdown scripts to suit their
project's needs [@Ziemann2023-kz].
Although we aimed to make this process as easy as possible, it requires a substantial time outlay, which
we estimate at 15-20 hours.
This is significantly longer than using a web-based tool that provides results in just a few
minutes, but we consider the time investment is well spent.
It ensures analytical reproducibility, assists with auditing and makes the process of reproduction
simple and fast (Box 1), which should aid the early identification of errors.
